### 2019-10-30 Writing/disparity_review.tex
Updated figures

\documentclass[12pt,letterpaper]{article}
\usepackage{natbib}

%Packages
\usepackage{xcolor}
\usepackage{color,soul}
\usepackage{pdflscape}
\usepackage{fixltx2e}
\usepackage{textcomp}
\usepackage{fullpage}
\usepackage{float}
\usepackage{latexsym}
\usepackage{url}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{array}
\usepackage[version=3]{mhchem}
\usepackage{ifthen}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{amsthm}
\usepackage{amstext}
\usepackage{enumerate}
\usepackage[osf]{mathpazo}
\usepackage{dcolumn}
\usepackage{lineno}
\usepackage{dcolumn}
\usepackage{hyphenat}
\usepackage[T1]{fontenc}
\usepackage{textcomp}
\newcolumntype{d}[1]{D{.}{.}{#1}}

\pagenumbering{arabic}

%Pagination style and stuff
\linespread{2}
\raggedright
\setlength{\parindent}{0.5in}
\setcounter{secnumdepth}{0} 
\renewcommand{\section}[1]{%
\bigskip
\begin{center}
\begin{Large}
\normalfont\scshape #1
\medskip
\end{Large}
\end{center}}
\renewcommand{\subsection}[1]{%
\bigskip
\begin{center}
\begin{large}
\normalfont\itshape #1
\end{large}
\end{center}}
\renewcommand{\subsubsection}[1]{%
\vspace{2ex}
\noindent
\textit{#1.}---}
\renewcommand{\tableofcontents}{}
%\bibpunct{(}{)}{;}{a}{}{,}

%---------------------------------------------
%
%       START
%
%---------------------------------------------

\begin{document}

%Running head
\begin{flushright}
Version dated: \today
\end{flushright}
\bigskip
\noindent RH: pest practices for disparity analysis.

\bigskip
\medskip
\begin{center}

\noindent{\Large \bf Analysing disparity in morphology data: best practice and future directions}

\bigskip

\noindent {\normalsize \sc Thomas Guillerme$^{1,*,+}$, Natalie Cooper$^{2,+}$, Stephen Brusatte$^3$, Katie Davis$^4$, Andrew Jackson$^5$, Sylvain Gerber$^6$, Anjali Goswami$^2$, Kevin Healy$^7$, Melanie Hopkins$^8$, Graeme Lloyd$^9$, Joseph O'Reilly$^{10}$, Abi Pate$^{10}$, Emilie Rayfield$^{10}$, Erin Saupe$^{11}$, Emma Sherratt$^{12}$, Graham Slater$^{13}$, Gavin H Thomas$^{14}$ and Philip Donoghue$^{10,+}$}\\

\noindent {\small \it 
$^1$School of Biological Sciences, University of Queensland, St. Lucia, Queensland, Australia.\\ 
$^2$Department of Life Sciences, Natural History Museum, London, Cromwell Road, London, SW7 5BD, UK.\\ 
$^{10}$Bristol}
\end{center}
\medskip
\noindent{+ These authors contributed equally to the manuscript.}
\noindent{*\bf Corresponding author.} \textit{guillert@tcd.ie}\\  
\vspace{1in}

%Line numbering
\modulolinenumbers[1]
\linenumbers

%---------------------------------------------
%
%       ABSTRACT
%
%---------------------------------------------

\newpage
\begin{abstract}

Variation in organisms' shape, ecology or behaviour is often more discrete rather than continuous.
One way to study this variation was pioneered in the 1990s by looking at the disparity of morphologies in the Cambrian.
Morphological disparity analysis - a term encompassing both the patterns and process of morphological evolution through time - became rapidly popular and lead to numerous publications in the last two decades.

Here, we suggest that this diversification of disparity analysis lead to two contrasted effects:
(1) the great diversity of methodological approaches to study disparity makes it sometimes hard to understand all caveats and pitfalls of such analysis - a problem that can lead to major misinterpretation of disparity patterns - and;
(2) in essence, disparity is a pattern of variation meaning that disparity analysis can be expanded beyond the study of morphology.

In this review, we propose a set of non-prescriptive guidelines that will help researchers to think about specific disparity analysis designs, with their caveats and their advantages.
Furthermore, we highlight potential avenues of expansion of disparity analysis both in terms of improving the existing methods for studying morphological disparity and in terms of expanding disparity analysis beyond morphology.

\end{abstract}

\noindent (Keywords: disparity)\\

\vspace{1.5in}

\newpage

%---------------------------------------------
%
%       INTRODUCTION
%
%---------------------------------------------
\section{Introduction}

Variation in the shape, form and function of organisms often follows discrete patterns rather than continuous gradients.
These patterns can be pictured as multidimensional spaces encompassing all the possible forms an organism can take, and where each individual region of this space is the result of a discrete combination of many variables.
One commonly used approach to analyse these discrete patterns, and to understand the processes that lead to them, is to analyse the disparity of the morphological characteristics of the group, often referred to as ``morphological disparity analysis''.
% The multidimensional space for mammalian carnivores, for example, encompasses all combinations of the traits of the order Carnivora, along with the ecological, life history and behavioural variables characteristic of meat eating mammals.
% Each individual carnivore species occupies a discrete region of this space related to its values for these traits.

In the seminal disparity papers from the late 1980s, disparity was used to investigate body plan evolution after the Cambrian explosion (REF)[Runnegar 1987, Gould, Foote, Webb, etc.].
Early disparity analyses focused on broad questions such as: How does diversity evolve? Are some body types more common than others? Can shape evolve in all ``directions'' (of the multidimensional space)?
Since the 1980s, a great deal of literature from a range of sub-fields in biology, and an equally large number of methods, have used the term ``disparity''.
Disparity analysis has been used to tackle a vast array of questions, for example in evo-devo (i.e. developmental constraints [CITE]; relationships between form and function [CITE]; evolution functional traits [CITE]; and scaling from micro- to macro-level processes [CITE]), in ecology and palaeoecology (i.e. intra-specific competition [CITE], niche occupancy[CITE]; speciation[CITE]; competition [CITE]; and macroecology/disparity in space[CITE]), and macroevolution (i.e. responses to extinction and extinction selectivity [CITE]; intra-specific competitive replacements [CITE]; and tempo and mode of evolution [CITE]).

The initial informal definition of disparity was ``multidimensional morphological dissimilarity at a macroevolutionary scale'' (REFS).
Currently, however, there is no consensus on whether we should restrict use of the term disparity to this initial definition, or whether it should be used as a broader umbrella term for studies of multivariate variation (REF Katie et al's paper somewhere here). 
Here we use a broader definition and suggest that disparity is a pattern of variation.
Throughout this review, we aim to highlight the diversity of ``disparity'' analyses and how they can aid our understanding of the structure of biodiversity.

% \subsection{What \textit{is} disparity?} %TG: maybe useful for box or paraphrapsing summaries.
% What \textit{is} disparity? In the seminal disparity papers from the late 80s, disparity was used in analyses of body plan evolution after the Cambrian explosion (REF).
% The initial informal definition of disparity was, therefore, ``multidimensional morphological dissimilarity at a macroevolutionary scale''.
% Currently, there is no consensus on whether we should restrict use of the term disparity to this initial definition, or whether it should be used as a broader umbrella term for studies of multivariate variation. 
% There is also debate about whether disparity refers to a pattern, a process that generates the pattern, or both. 
% Even if disparity only describes patterns of variation, the scientific questions of disparity analyses generally use these patterns as a proxy for explaining the biological processes underlying them.
% The original Cambrian explosion papers, for example, use disparity to describe both the pattern of high dissimilarity in body plans among groups, and the process of explosive radiation that generated them (as championed by Gould; REF).

% Restricting disparity to its original definition can be advantageous for advances in this field but broadening it allows more interdisciplinary discussion.
% In recent years, disparity has been used to investigate ecological processes, disconnected from evolution (e.g. function [CITE], community structure[CITE], competition[CITE]).
% Thus studying disparity could not only inform us on macroevolutionary patterns and processes but also something about environments and ecosystem functioning (with the advantages of easily allowing palaeocological and palaeoenvironemental perspectives).

\section{Best practices for disparity analysis}

This diversity (and disparity) of methods to analysis disparity provides biologists with a powerful and comprehensive toolkit for understanding the evolution of biological diversity.
However, as with any great power, it comes with great responsibility. 
With the expansion of methodological approaches in the last decade, we identify three main issues that could reduce the power of disparity analyses if they are not tackled.
First, many recent studies have lost focus; they are no longer hypothesis-driven, instead they are undertaken to characterise biological variation for its own sake.
Second, even where there is a clear question being explored, the data being used for disparity analyses is inappropriate and often driven by the availability of large datasets rather than the question being asked.
Third, the methods being used to analyse these datasets may not always be appropriate for the question and/or data at hand.
Here we deal with some of these issues and present best practice guidelines to allow researchers to optimise their disparity analysis.

\section{What is the question?}
Why perform a disparity analysis?
This seems like it should be obvious, but we stress its crucial importance.
The last decade has seen a vast array of new methods being published and implemented in freely available software, making disparity analyses easy to perform \citep{bouxin2005ginkgo,oksanen2007vegan,geiger2008,zelditch2012geometric,adams2013geomorph,Claddis,dispRityv02,adams2017geometric}.
However, this (consciously or not) pushes authors towards performing disparity analyses simply because of their ease of implementation, rather than stemming from a question driven approach.
Before rushing into a disparity analysis, one should consider some basic caveats.
For example, one major thing to consider is circularity: if the aim is to explore the variation in a dataset, disparity analysis is perfectly valid.
However, if this exploratory analysis reveals distinct groups, it would be circular to then test for differences among those groups as if this had been the hypothesis all along [CITE stats 1.0].
This can easily be avoided by using a subset of the data for preliminary data exploration, and then using the full dataset to test for differences among any groups that emerge.
Clearly framing questions and their null expectations will help in designing better disparity analyses.

\section{Collect the appropriate data}
Once one has a clear question, it is important to collect the appropriate data for that question to avoid some major caveats.
For example, if a study focuses on the ecological adaptive radiation of a group, ecological traits are crucial and should be preferred over other kinds of traits, such as those used for species delimitation.
Tangentially, more data are not necessarily better, especially when the data were originally collected for different questions (e.g. cladistic data - see below).

\subsection{Data types}
Morphological disparity analysis can be based on three main types of data:
\begin{enumerate}
    \item discrete data that can either describe a feature (often referred to as ``cladistic''; e.g. the presence or absence of legs, the colour of a tail, etc. [CITE]) or count the number of features (e.g. number of digits, etc. [CITE]);
    \item continuous data that can be either simple measurements (e.g. body length, etc. [CITE]) or more abstracted measures (e.g. geometric morphometrics, mathematical contour description, etc. [CITE]);
    \item any mixture of discrete and continuous data (e.g. ratios [CITE]).
\end{enumerate}
None of these types of data are fundamentally better but each will be more appropriate for specific questions.
For example, when investigating variation within bat wing shapes, both homologous landmarks (for geometric morphometric analysis) or continuous measurements of bones may be appropriate.
On the other hand, if the question focuses on the convergence between bat and bird wings, the homology of the traits to measure is more complex to define and the outline shape of the wings might more readily answer the question.
This again illustrates the importance of framing the question.
In the above example, if the question is whether the aerodynamic properties of wings vary within bats or between bats and birds, the traits collected should reflect these aerodynamic properties (e.g. wingspan, aspect ratio, etc.).
Conversely if the question focuses on phylogenetic convergence aspects between different types of species with active flight, it might be more pertinent to use evolutionary traits that may have facilitated flight within and between both groups (e.g. digit length, integumentary system, etc.).

\subsection{How much data should be collected?}
Regardless of the type of data, another important consideration is how much data to collect, in terms of numbers of specimens and traits.
\begin{itemize}
    \item \textit{Taxonomic scope.}
    Morphological disparity has been studied at both inter-[CITE] and intra-specific levels [CITE].
    Taxonomic scope will depend on the question being investigated, and what type of variation is being analysed.
    For example, when studying skull shape plasticity in canids, it is essential to include intra-specific variation to capture the full variety of skull shapes, particularly that seen in domestic dogs (CITE).
    Conversely, if a study focuses on canid skull variation through time, interspecific data is most appropriate.
    % intra-specific variation will not have major impact (up until the Holocene). %http://www.journals.uchicago.edu/doi/abs/10.1086/650372
    \item \textit{Sample size.}
    It is important to ensure that the data is collected in a way that will not bias subsequent statistical analysis.
    For example, if the aim is to compare disparity among groups, one should collect, where possible, a similar amount of data for each group.
    When this is not possible (e.g. for palaeontological data), one should use statistical approaches to correct or assess biases due to data collection (e.g. rarefaction [CITE]).
    Equally, one should ensure that enough data is available so that the methods have power to detect the patterns being tested.
    \end{itemize}

Of course in practice some of these caveats are unavoidable, especially when dealing with the vertebrate fossil record.

\subsection{Recycling data}
A common practice in morphological disparity analysis is to recycle data from other studies which can introduce additional problems.
One classic case is using ``cladistic'' data in palaeobiology.
These data are usually collected for phylogenetic analysis and are thus designed to maximise differences among taxa to recover a clear phylogenetic signal.
It can therefore artificially lead to maximised disparity among taxa [Foote - crinoids non-cladistic paper].
A relatively simple way to reduce this specific bias is to select the most pertinent characters for the question from the dataset, rather than using all of the data.
For example, in a study of the disparity of dietary niches, removing non-dietary characters is a good approach.
However, it is important to remember that there is no one-to-one mapping between form and function, and that some characters can have multiple functions (e.g. limb traits can be used for both locomotion and feeding).

\subsection{Other data considerations}
There are many other considerations when choosing the appropriate data to use. 
It is essential to ensure that disparity patterns can be linked to general processes rather than being simply the reflection of idiosyncrasies in the data.
For example, are the observed differences between two groups related to factors unrelated to group designation such as sexual dimorphism?
Are the taxa/groups being compared fairly consistent and truly comparable?
In addition, data can be missing, non-overlapping, hierarchical (e.g. inapplicable), ambiguous, polymorphic or correlated [String of morpho data problems cite].
It is important to ensure that any patterns detected are not merely reflections of these features of the data.

Ultimately, data collection is often constrained by the time available, financial considerations and other unavoidable issues, however, these caveats should be kept in mind during the analyses and subsequent interpretation of the results.

\section{Use the appropriate methods} 
In the vast majority of morphological disparity analyses, the methodological approach will be multidimensional.
Multidimensional analysis can be a powerful tool for analysing complex datasets but also comes with many mathematical caveats and limitations.
The first step in any statistical analysis should be to visualise the data to understand its distribution and limitations.
Unfortunately, we can only perceive a maximum of three dimensions at the same time.
Therefore one of the main challenges in multidimensional analysis is how to reduce the number of dimensions, both for visualisation and analysis (see box \ref{box_ordination}).

\subsection{Visualising the multidimensional problem}
\label{visualisation}
It is important to remember that visual interpretations of multidimensional data can easily be misleading.
For example, it is incorrect to assume that because some groups overlap on a bivariate or three dimensional plot, that they overlap across the whole multidimensional space (especially if the space is not Euclidean).
Conversely, when two groups do not overlap on a two or three dimensional plot, this does not mean they never overlap in the multidimensional space.
Furthermore, multidimensional spaces are not necessary Euclidean.
Such non-Euclidean spaces often have non-intuitive properties, for example straight lines in some dimensions are not straight in the space and, even less intuitively, the distance between points A and B might not be equal to the distance between points B and A [CITE Sylvain].

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.5\textwidth]{Figures/dimensionsOverlap.pdf}
\caption{\small{Two groups that overlap in some dimensions do not necessary overlap in all dimensions. 
Conversely, two groups that do not overlap in some dimensions can overlap in other dimensions.
In this example for the groups A and B to truly overlap, they must overlap in all three dimensions displayed here.
Although these groups overlap in the two first dimensions (panel 1), they do not overlap in dimensions 1 and 3 thus do not overlap in the multidimensional space at all.
The inverse is of course true: if two groups don't overlap in two dimensions (D and C in panel 1), they do not overlap in the multidimensional space, even if they overlap in any other combination of dimensions (dimensions 1 and 3 in panel 2).}} 
\label{Fig:RF_results_best}
\end{figure}

It is therefore vital to be clear about what is represented in multidimensional space visualisations.
In fact, it can be easy (intentionally or not) to mislead a reader into over/under interpreting a pattern if all the key properties of the space are not present in the graph.
We suggest visualisations should include the following:
(1) the percentage of the variation explained by each axis of the plot (i.e. is the majority - or all - the variation in the data present in the plot or only a small fraction?);
(2) the scale of the axes (ideally, if the space is orthogonal, it is important to keep the same scale across the axes);
and (3) the error associated with each point, whether this is in terms of measurement error (on ordination plots for example) or some other measure of standard variation (in further analysis, e.g. disparity-through-time plots).
Finally, it is important to remember that for any dimensionality reduction of a multidimensional problem, the visual result will be at best a distortion of the measured reality, at worst, it will be misleading (see box \ref{box_ordination}).

\subsection{[Box]To ordinate or not to ordinate; that is the (multidimensional) question?}
\label{box_ordination}
Reducing the dimensionality of data through ordination can be advantageous for plotting and visualising the data, and can reveal properties of the morphospace not captured by disparity metrics (see Metrics section \ref{metrics}).
Additionally, after ordinating the data it is possible to further reduce the number of dimensions by only selecting a certain number of axes (e.g. selecting only the axes that describe up to 95\% of the variation).
In the case of geometric morphometric data, ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions.
This has clear advantages for interpreting the results.
For example, the axes will represent gradients of biological variation (e.g. elongation and flattening of the beak REF).

>>>>

RH: disparate disparity analyses.

\textbf{The disparate ways to analyse morphological disparity}

Thomas
Guillerme\includegraphics[width=0.4902in,height=0.1807in]{./ObjectReplacements/Object 2},
Natalie
Cooper\includegraphics[width=0.3984in,height=0.1807in]{./ObjectReplacements/Object 4},
Stephen
Brusatte\includegraphics[width=0.1937in,height=0.1807in]{./ObjectReplacements/Object 6},
Katie
Davis\includegraphics[width=0.1992in,height=0.1807in]{./ObjectReplacements/Object 8},
Andrew L
Jackson\includegraphics[width=0.1937in,height=0.1807in]{./ObjectReplacements/Object 10},
Sylvain
Gerber\includegraphics[width=0.1953in,height=0.1807in]{./ObjectReplacements/Object 12},
Anjali
Goswami\includegraphics[width=0.1929in,height=0.1807in]{./ObjectReplacements/Object 14},
Kevin
Healy\includegraphics[width=0.1953in,height=0.1807in]{./ObjectReplacements/Object 16},
Melanie
Hopkins\includegraphics[width=0.1953in,height=0.1807in]{./ObjectReplacements/Object 18},
Graeme
Lloyd\includegraphics[width=0.1953in,height=0.1807in]{./ObjectReplacements/Object 20},
Joseph
O'Reilly\includegraphics[width=0.2437in,height=0.1807in]{./ObjectReplacements/Object 22},
Abi
Pate\includegraphics[width=0.2437in,height=0.1807in]{./ObjectReplacements/Object 24},
Emilie
Rayfield\includegraphics[width=0.2437in,height=0.1807in]{./ObjectReplacements/Object 26},
Erin
Saupe\includegraphics[width=0.239in,height=0.1807in]{./ObjectReplacements/Object 28},
Emma
Sherratt\includegraphics[width=0.2425in,height=0.1807in]{./ObjectReplacements/Object 30},
Graham
Slater\includegraphics[width=0.2437in,height=0.1807in]{./ObjectReplacements/Object 32},
Gavin H
Thomas\includegraphics[width=0.2465in,height=0.1807in]{./ObjectReplacements/Object 34}
and Philip
Donoghue\includegraphics[width=0.2437in,height=0.1807in]{./ObjectReplacements/Object 36}\\
\includegraphics[width=0.1929in,height=0.1807in]{./ObjectReplacements/Object 38}\emph{School
of Biological Sciences, University of Queensland, St. Lucia, Queensland,
Australia.}\\
\includegraphics[width=0.1929in,height=0.1807in]{./ObjectReplacements/Object 40}\emph{Department
of Life Sciences, Natural History Museum, London, Cromwell Road, London,
SW7 5BD, UK.}\\
\includegraphics[width=0.2437in,height=0.1807in]{./ObjectReplacements/Object 42}\emph{Bristol}

+ These authors contributed equally to the manuscript.

*\textbf{Corresponding author: }\emph{guillert@tcd.ie}\\

\textbf{NOTE TO ALL THE CO-AUTHORS:}

\textbf{-Please write in Google docs using }\emph{\textbf{Suggesting
mode}}\textbf{ only }(upper right corner of your screen)

\textbf{-The word limit for this is 5000 words so we can't add any large
sections.}

\textbf{-There is no reference limit per se but please don't add too
many. I will sort out the references later so don't correct the
formatting of them here.}

-I suggest you add your references using the \textbf{Paperpile} addon
(go in Add-ons menu \textgreater{} Get Add-ons\ldots{} \textgreater{}
Search for Paperpile \textgreater{} Add/Install; then go in Add-ons
menus \textgreater{} Paperpile \textgreater{} Manage citation). If you
don't use Paperpile, please don't add references as just (``Someone et
al. 2000'' but add also a link to the correct Google Scholar entry in
comments).

\textbf{-Add your affiliation in the author list above}. The
affiliations count in the word count so please 1) keep it short (max 1
per person!) and 2) try to use one that's already in there. Also add any
initials I've missed in your names.

- If in doubt, check formatting here:
\href{https://royalsociety.org/journals/authors/author-guidelines/}{\emph{https://royalsociety.org/journals/authors/author-guidelines/}}

Thanks!

\textbf{Abstract (200 words max)}

Morphological disparity analysis is used to characterize and investigate
discrete patterns of variation in the shape, form, function, and the
ecology of organisms. Since their conception in the 1980s, these methods
have been used to ask a variety of questions about how and why
morphological diversity evolves, and the ecological consequences of
these changes. There are many methodological approaches for studying
disparity, however, all of these come with caveats and pitfalls that
can, if ignored, lead to misinterpretations of disparity patterns. Here
we review morphological disparity methods and their caveats, and provide
best practice guidelines for their application and interpretation. We
discuss the data required for disparity studies, as well as
methodological challenges, including multidimensionality, ordination,
metric choice, hypothesis testing, and ancestral state estimations.
There is no ``one-size-fits-all'' approach to characterizing and
investigating the evolution of morphological disparity, and the
available tools should always be used in the context of a specific
biological question that will determine data and method selection at
every stage of the analysis.

\textbf{Keywords}: multidimensionality, palaeobiology, ecology\\

1. Introduction

Variation in the shape, form and function of organisms often exhibits a
pattern of discrete clusters rather than continuous gradients. These
patterns can be pictured as multidimensional spaces encompassing all the
possible forms an organism can take, and where each individual region of
this space is the result of a unique combination of many variables. One
commonly used approach to analyse these patterns and to understand the
processes that lead to them, is to analyse the morphological diversity
of a group, often referred to as a ``morphological disparity analysis''
\href{https://paperpile.com/c/sTGYvp/0y4V}{(Deline et al. 2018)}.

The initial informal definition of morphological disparity was
``multidimensional morphological dissimilarity at a macroevolutionary
scale'' (Runnegar 1987,
\href{https://paperpile.com/c/sTGYvp/Uns3}{(Gould 2000)}). In the
seminal disparity papers of the late 1980s, disparity was used to
investigate body plan evolution after the Cambrian explosion
\href{https://paperpile.com/c/sTGYvp/CiPy}{(Briggs, Fortey, and Wills
1992)}\href{https://paperpile.com/c/sTGYvp/CidX+6tNm+oenu+TtGs}{(Gould
1991; Valentine 1986; M. Foote and Gould 1992; Morris 1989)};
\href{https://paperpile.com/c/sTGYvp/CiPy}{(Briggs, Fortey, and Wills
1992)};. Early disparity analyses focused on broad questions such as:
How does diversity evolve? Are some phenotypes more common than others?
and Can anatomical shapes evolve in all ``directions'' of the
multidimensional space
\href{https://paperpile.com/c/sTGYvp/yqPw+fTJ3}{(Mike Foote 1997,
1995)}? Since the 1990s, a great deal of literature from a range of
sub-fields in biology and palaeontology, and an equally large number of
methods, have used ``disparity'' as a broader umbrella term for studies
of multivariate variation. Although most influential reviews have
treated disparity as a pattern
\href{https://paperpile.com/c/sTGYvp/yqPw+nFf7+vTHS}{(Mike Foote 1997;
Wills 2001; Melanie J. Hopkins and Gerber 2017)}, there is currently no
consensus on whether to restrict use of the term disparity to its
original macroevolutionary-scale definition, or whether it can refer to
any study of multivariate variation. Here we use a broader definition
and suggest that morphological disparity is a \emph{pattern of variation
}in phenotype. Disparity analysis has been used to tackle a vast array
of questions. We identify four main classes of disparity analysis as
follows.

(1) \emph{Descriptive disparity} \emph{analyses. }The seminal studies of
disparity described the shapes of organisms and how they differed among
groups \href{https://paperpile.com/c/sTGYvp/fTJ3+eZ3F}{(Mike Foote 1995;
Wills, Briggs, and Fortey 1994)}. These consist of descriptive
multidimensionality analyses of the patterns in the diversity of
morphological traits, addressing questions such as: why are some
morphological traits combinations more common than others and what are
the properties of the resulting morphospace?
\href{https://paperpile.com/c/sTGYvp/fTJ3+I0Ic+QVvv}{(Mike Foote 1995;
Raup 1961; Gerber 2017)}.

(2) \emph{Disparity-through-time} \emph{analyses. }This approach
investigates how the shapes of organisms have changed through time,
focussing on the disparity of taxa in particular time bins or slices.
This approach has been widely used in palaeobiology to answer a range of
macroevolutionary questions
\href{https://paperpile.com/c/sTGYvp/PbSx+khc9+xxh5}{(Close et al. 2015;
Cisneros and Ruta 2010; Hughes, Gerber, and Wills 2013)}, such as: how
does disparity accumulate over the history of a clade
\href{https://paperpile.com/c/sTGYvp/ekU4+s33b}{(Guillerme and Cooper
2018; Wright 2017)}, and how does disparity change across mass
extinction events \href{https://paperpile.com/c/sTGYvp/EETc}{(Friedman
2010)}?

(3) \emph{Disparity and diversity. }Morphological disparity provides
another perspective on biodiversity; high morphological disparity
represents a high diversity of morphologies (i.e. shapes or body plans)
and is, presumably, associated with high levels of ecological and
functional diversity (see below). This makes disparity a more useful
measure of diversity than merely species richness. Indeed, most studies
that have investigated disparity and taxonomic diversity support an
effective decoupling
\href{https://paperpile.com/c/sTGYvp/2tbJ+geAO+hea5+aVVj}{(Fortey,
Briggs, and Wills 1996; Ruta et al. 2013; M. J. Hopkins 2013; Moyne and
Neige 2007)} - though this result breaks down at extremes of diversity;
extremely low diversity tends to yield low disparity and \emph{vice
versa}. This approach has been used to investigate whether some groups
are more successful than other in their exploration of new evolutionary
strategies @@@.

(4) \emph{Disparity and ecology. }The disparity of a group can be used
as a proxy for either the functional role it plays within an ecosystem
or its ecological niche. These analyses assume that groups with high
disparity are also likely to be functionally and ecologically diverse,
and that groups found in similar regions of shape space will have
similar functional and ecological roles
\href{https://paperpile.com/c/sTGYvp/EETc+tSIy+qjj9}{(Friedman 2010;
Pierce, Angielczyk, and Rayfield 2008; P. S. L. Anderson et al. 2011)}.
Note however, that the links between form and function are not always
clear cut; traits can be linked to multiple functions and multiple
functions can be linked to a single trait
\href{https://paperpile.com/c/sTGYvp/Ejzr}{\emph{(Wainwright et al.
2005)}}. This approach is used to investigate competitive replacements
\href{https://paperpile.com/c/sTGYvp/EeC8}{(Stephen L. Brusatte et al.
2008)} and changes in ecosystem function during and after mass
extinctions \href{https://paperpile.com/c/sTGYvp/EETc}{(Friedman 2010)}.
It is particularly common in paleobiology where it is not possible to
observe directly the ecological or functional characteristics of extinct
species. \href{https://paperpile.com/c/sTGYvp/Ejzr}{(Wainwright et al.
2005)}

Of course, these four broad categories are not independent and many
studies use a combination of approaches
\href{https://paperpile.com/c/sTGYvp/EeC8}{(Stephen L. Brusatte et al.
2008)}. Furthermore, disparity analysis is continuously expanding into
areas. While this is exciting for the field, there are several issues
that may reduce the utility of disparity analyses if they are not
tackled. Here we review these issues and present best practice
guidelines to allow researchers to optimise their analyses. We first
discuss the appropriate data required for disparity analyses, then
review various challenging methodological aspects of disparity analyses.
Throughout it is important to remember that these tools should always be
used in the context of a specific scientific question, as this will
drive data and methodological choices at every stage of the process.

\includegraphics[width=6.5in,height=4.9445in]{Pictures/10000201000002EE0000023BE5B78D4C4B79F709.png}

\emph{Figure disparities {[}figure\_disparities.pdf{]}: The four main
types of disparity analysis. Descriptive disparity analyses focus on
describing the properties of the morphospace ; Disparity-through-time
analyses investigate the evolution of the morphospace through time
including the effect of extinction events; Disparity and diversity
analysis use disparity as a proxy for species diversity; Disparity and
ecology analyses use disparity as a proxy for the ecological or
functional role of a group. These categories are not independent and
many analyses will cover more than one.}

2. Data in Disparity Analyses

Regardless of the precise approach being used, disparity analyses are
based on morphological traits. It is important, therefore, to consider
which traits to use in the context of the specific question under
investigation. Traits for disparity analyses come from three main types
of data: (1) discrete ``cladistic'' characters, i.e coding the
absence-or-presence of features or a discrete characteristic of a trait
\href{https://paperpile.com/c/sTGYvp/PbSx}{(Close et al. 2015)}); (2)
continuous measurements of features
\href{https://paperpile.com/c/sTGYvp/qjj9}{(P. S. L. Anderson et al.
2011)}; or (3) more sophisticated mathematical measurements from
geometric morphometric data (e.g. Procrustes superimpositions
\href{https://paperpile.com/c/sTGYvp/RjqE}{(Cooney et al. 2017)}, or
Fourier contours \href{https://paperpile.com/c/sTGYvp/ZEDR}{(Spriggs et
al. 2018)}). None of these are fundamentally better, however, some
questions and methodological pipelines are more appropriate with
specific types of data.

When investigating variation within bat wing shapes, for example, both
homologous landmarks from geometric morphometric analysis and continuous
measurements of bones may be appropriate to capture patterns of wing
variation. However, if the question focuses on the convergence between
bat and bird wings, the homology of the traits to measure is more
complex to define and the outline shape of the wings might more readily
answer the question from a mechanical perspective. Similarly, if the
focus is on wing function, i.e. whether the aerodynamic properties of
wings vary within bats or between bats and birds, the traits collected
should reflect these aerodynamic properties (e.g. wingspan, aspect
ratio, etc.). Whereas if the focus is on convergence between different
bats and birds, it would be preferable to use traits that have
facilitated flight in both groups (e.g. digit length, integumentary
system, etc.). Where there is any doubt about the appropriate traits to
choose, we advise using different kinds of data for the same feature to
determine whether it captures the same disparity.

This dicussion assumes that researchers are collecting their own data
for disparity analyses, but often this is not the case. It is common,
especially in disparity-through-time analyses, to recycle discrete
character data collected for use in phylogenetic inference (e.g. REFs
that have done this). This has the advantage of being quick, cheap and
efficient, but tends to artificially increase disparity between
phylogenetically distinct groups because phylogenetic characters are
collected specifically for the purpose of discriminating groups
\href{https://paperpile.com/c/sTGYvp/fTJ3}{(Mike Foote 1995)}. This
needs to be considered when interpreting results, especially as when new
clades appear, the synapomorphies of that clade will naturally lead to
an apparent shift or increase in disparity. Similarly, data used for
phylogenetic reconstruction is likely to be highly skewed towards easily
available and taxonomically relevant traits, for example teeth in
mammals. If your disparity question is related to feeding ecology or
similar then this may be appropriate, but if the focus is on dispersal
ability then it would be sensible to select only postcranial traits. It
is often tempting to use all of a large dataset to increase the power of
your analyses, but if the data are not suitable for the question at hand
then the results may be meaningless (although ssome studies have suggest
this may not be a major issue;
\href{https://paperpile.com/c/sTGYvp/ZqML+Qrba}{(Zou and Zhang 2016;
Dávalos et al. 2014)}
\href{https://paperpile.com/c/sTGYvp/xLdm}{(Melanie J. Hopkins 2017)}).

Finally, trait data suffers from the same shortcomings as most data --
data can be missing, non-overlapping, hierarchical, inapplicable,
ambiguous, polymorphic, correlated, or have insufficient sample size
\href{https://paperpile.com/c/sTGYvp/yO2t}{(Palci and Lee 2018)},
\href{https://paperpile.com/c/sTGYvp/Yrbg}{(Brazeau, Guillerme, and
Smith 2017)} {[}String of morpho data problems cite{]}. Biological
phenomena such as allometry and sexual dimorphism may also influence
trait data. More practically, data collection is constrained by the time
and money available, making collating a ``perfect'' dataset difficult.
We suggest that trait data are collected with the question in mind and
considering all of the caveats above, but where issues are unavoidable
at the very least these caveats must be considered during the analyses
and subsequent interpretation of the results.

\emph{Figure data {[}figure\_data.pdf{]}: major routes to obtain
disparity data. Data can be collected as discrete trait observations
(e.g. presence or absence data) or as continuous data. Continuous data
can be collected by different methods: linear measurements, landmark
coordinates or contours (curves). These measurement can then in turn be
mathematically transformed (logarithm, scaling, Procrustes
Superimposition, Elliptic Fourier, etc.). Regardless of the method, this
data collection results in a trait matrix where the observed traits
constitute columns and the studied elements (e.g. taxa) the rows.}

\includegraphics[width=6.5in,height=6.5in]{Pictures/10000201000007D0000007D0F8A72A6590C667D3.png}

3. Disparity Methods

Once suitable trait data have been collected, we need to consider the
disparity analysis. Below we address several key aspects including (i)
the difficulty of dealing with multidimensional data; (ii) caveats
associated with using ordination techniques to reduce the dimensionality
of the data; (iii) various metrics used to summarise the disparity of a
group; (iv) caveats related to methods that exist for hypothesis testing
within a disparity analysis framework; and (v) the specific case of
ancestral state estimations in disparity-through-time analyses. As noted
above, these tools should always be used in the context of a specific
question, as this will determine which methods are most appropriate.

\protect\hypertarget{anchor}{}{}3.1. Multidimensional mayhem
